Certified Knowledge Compilation \\ with Application to Verified Model Counting

### 2023-03-02 doc/mc-paper.tex
Tweaked figure

\newcommand{\varset}{X}
\newcommand{\exvarset}{Z}
\newcommand{\dependencyset}{{\cal D}}
\newcommand{\litset}{{\cal L}}
\newcommand{\ring}{{\cal R}}
\newcommand{\dset}{{\cal A}}
\newcommand{\rep}{\textbf{R}}
\newcommand{\radd}{+}
\newcommand{\rmul}{\times}
\newcommand{\addident}{\textbf{0}}
\newcommand{\mulident}{\textbf{1}}

@@

%\authorrunning{R. E. Bryant, et al} %TODO mandatory. First: Use abbreviated first/middle names. Second (only in severe cases): Use first author plus 'et al.'

@@

  \centering{
\begin{minipage}{0.58\textwidth}
(A)  Input Formula\\[1.2ex]
\begin{tabular}{lll}
\toprule
\makebox[5mm]{ID} & \makebox[15mm]{Literals} & \\
\midrule
1 & \texttt{-1 3 -4} & \texttt{0} \\
2 & \texttt{-1 -3 4} & \texttt{0} \\
3 & \texttt{3 -4} & \texttt{0}\\
4 & \texttt{1 -3 4} & \texttt{0} \\
5 & \texttt{-1 -2} & \texttt{0} \\
\bottomrule
\end{tabular}
\\[1.8ex]
(B) CRAT Declaration of POG\\[1.2ex]
\begin{tabular}{llll}
\toprule
\makebox[5mm]{ID} & \multicolumn{2}{l}{Proof line} & Explanation \\
\midrule
6 & \texttt{p 5 -3 -4} & \texttt{0} & $p_5 = \obar{x}_3 \pand \obar{x}_4$ \\
9 & \texttt{p 6 3 4} & \texttt{0} & $p_6 = x_3 \pand x_4$ \\
12 & \texttt{s 7 5 6 7 10} & \texttt{0} & $s_7 = p_5 \por p_6$ \\
15 & \texttt{p 8 -1 7} & \texttt{0} & $p_8 = \obar{x}_1 \pand s_7$ \\ 
18 & \texttt{p 9 1 -2 7} & \texttt{0} & $p_9 = x_1 \pand \obar{x}_2 \pand s_7$ \\
22 & \texttt{s 10 8 9 16 19} & \texttt{0} & $s_{10} = p_8 \por p_9$ \\
 & \texttt{r 10} && Root $r = s_{10}$\\
\bottomrule
\end{tabular}
\end{minipage}
\begin{minipage}{0.40\textwidth}
(C) POG Visualization\\
\input{dd/eg4}
\end{minipage}
}
\caption{Example CNF Formula and POG Representation}
\label{fig:eg4:pog}
\end{figure}

Figure \ref{fig:eg4:pog} illustrates an example formula and shows how
the CRAT file declares its POG representation.  The input formula (A)
consists of five clauses over variables $x_1$, $x_2$, $x_3$, and
$x_4$.  The generated POG has six nonterminal nodes representing four
conjunctions and two disjunctions.  We name these by the node type (product or sum) and the extension variable.
The first part of the CRAT file
(B) declares these nodes using clause IDs that increment by three or
four, depending on whether the node has two children or three.
The last two nonzero values in the sum declarations are hints providing the required mutual exclusion proof.

\begin{figure}
\begin{minipage}{0.49\textwidth}
(A) Implicit clauses\\[1.2ex]
\begin{tabular}{llll}
\toprule
\makebox[5mm]{ID} & \multicolumn{2}{l}{Proof line} & Explanation \\
\midrule
\texttt{6} & \texttt{5 3 4} & \texttt{0} & Define $p_5$ \\
\texttt{7} & \texttt{-5 -3} & \texttt{0} & \\
\texttt{8} & \texttt{-5 -4} & \texttt{0} & \\
\midrule
\texttt{9} & \texttt{6 -3 -4} & \texttt{0} & Define $p_6$ \\
\texttt{10} & \texttt{-6 3} & \texttt{0} & \\
\texttt{11} & \texttt{-6 4} & \texttt{0} & \\
\midrule
\texttt{12} & \texttt{-7 5 6} & \texttt{0} & Define $s_7$ \\
\texttt{13} & \texttt{7 -5} & \texttt{0} & \\
\texttt{14} & \texttt{7 -6} & \texttt{0} & \\
\midrule
\texttt{15} & \texttt{8 1 -7} & \texttt{0} & Define $p_8$ \\
\texttt{16} & \texttt{-8 -1} & \texttt{0} & \\
\texttt{17} & \texttt{-8 7} & \texttt{0} & \\
\midrule
\texttt{18} & \texttt{9 -1 2 -7} & \texttt{0} & Define $p_9$ \\
\texttt{19} & \texttt{-9 1} & \texttt{0} & \\
\texttt{20} & \texttt{-9 -2} & \texttt{0} & \\
\texttt{21} & \texttt{-9 7} & \texttt{0} & \\
\midrule
\texttt{22} & \texttt{-10 8 9} & \texttt{0} & Define $s_{10}$ \\
\texttt{23} & \texttt{10 -8} & \texttt{0} & \\
\texttt{24} & \texttt{10 -9} & \texttt{0} & \\
\bottomrule
\end{tabular}
\end{minipage}
\begin{minipage}{0.49\textwidth}
(B) Assertions\\[1.2ex]
\begin{tabular}{llllll}
\toprule
\makebox[5mm]{ID} & \multicolumn{2}{l}{Proof line} & \multicolumn{2}{l}{Hints} & Explanation \\
\midrule
\texttt{25} & \texttt{5 1 3} & \texttt{0} & \texttt{3 6} & \texttt{0} & $\obar{x}_1 \land \obar{x}_3 \rightarrow p_5$ \\
\texttt{26} & \texttt{6 1 -3} & \texttt{0} & \texttt{4 9} & \texttt{0} & $\obar{x}_1 \land x_3 \rightarrow p_6$ \\
\texttt{27} & \texttt{3 7 1} & \texttt{0} & \texttt{13 25} & \texttt{0} & $\obar{x}_3 \land \obar{x}_1 \rightarrow s_7$  \\
\texttt{28} & \texttt{7 1} & \texttt{0} & \texttt{27 14 26} & \texttt{0} & $\obar{x}_1 \rightarrow s_7$  \\
\texttt{29} & \texttt{8 1} & \texttt{0} & \texttt{28 15} & \texttt{0} & $\obar{x}_1 \rightarrow p_8$  \\
\texttt{30} & \texttt{5 -1 3} & \texttt{0} & \texttt{1 6} & \texttt{0} & $x_1 \land \obar{x}_3 \rightarrow p_5$ \\
\texttt{31} & \texttt{6 -1 -3} & \texttt{0} & \texttt{2 9} & \texttt{0} & $x_1 \land x_3 \rightarrow p_6$ \\
\texttt{32} & \texttt{3 7 -1} & \texttt{0} & \texttt{13 30} & \texttt{0} & $\obar{x}_3 \land x_1 \rightarrow s_7$  \\
\texttt{33} & \texttt{7 -1} & \texttt{0} & \texttt{32 14 31} & \texttt{0} & $x_1 \rightarrow s_7$  \\
\texttt{34} & \texttt{9 -1} & \texttt{0} & \texttt{5 33 18} & \texttt{0} & $x_1 \rightarrow p_9$  \\
\texttt{35} & \texttt{1 10} & \texttt{0} & \texttt{23 29} & \texttt{0} & $\obar{x}_1 \rightarrow s_{10}$  \\
\texttt{36} & \texttt{10} & \texttt{0} & \texttt{35 24 34} & \texttt{0} & $s_{10}$ \\
\bottomrule
\end{tabular}
\\[1.2ex]
(C) Input clause deletions\\[1.2ex]
\begin{tabular}{lllll}
  \toprule
 \multicolumn{3}{l}{Proof line} & Explanation\\
\midrule
 \texttt{dc 1} & \texttt{36 8 10 12 16 21 22} & \texttt{0} & Delete clause 1 \\
 \texttt{dc 2} & \texttt{36 7 11 12 16 21 22} & \texttt{0} & Delete clause 2 \\
 \texttt{dc 3} & \texttt{36 8 10 12 17 19 22} & \texttt{0} & Delete clause 3 \\
 \texttt{dc 4} & \texttt{36 7 11 12 17 19 22} & \texttt{0} & Delete clause 4 \\
 \texttt{dc 5} & \texttt{36 16 20 22} & \texttt{0} &  Delete clause 5 \\
\end{tabular}
\end{minipage}
\caption{Defining Clauses (A), Assertions (B), and Input Clause Deletions (C) for Example Formula}
\label{fig:eg4:proof}
\end{figure}

We step through a detailed example to provide a better understanding of the CRAT proof framework.
Figure
\ref{fig:eg4:proof}(A) shows the defining clauses that are implicitly
defined by the POG operation declarations.  These do not appear in the
CRAT file.  Referring back to the declarations of the sum nodes in
Figure \ref{fig:eg4:pog}, we can see that the declaration of node
$s_7$ had clause IDs 7 and 10 as hints.  We can see in Figure
\ref{fig:eg4:proof}(A) that these two clauses resolve to the clause
$\obar{p}_5 \lor \obar{p}_6$, showing that the two children of $s_7$
have disjoint models.  Similarly, node $s_{10}$ is declared as having
clause IDs 16 and 19 as hints.  These resolve to the clause
$\obar{p}_8 \lor \obar{p}_9$, showing that the two children of
$s_{10}$ have disjoint models.

Figure \ref{fig:eg4:proof}(B) provides the sequence of assertions
leading to unit clause 36, consisting of the literal $s_{10}$.  This clause indicates that $s_{10}$ is implied by the input clauses, i.e.,
any assignment $\assign$
satisfying the input clauses must extend to one with $\assign(s_{10}) = 1$.
Working backward, we can see that
clause 29 indicates that node $p_8$ will be implied by the input
clauses when $\assign(x_1) = 0$, while clause 34 indicates that node $p_9$ will
be implied by the input clauses when $\assign(x_1) = 1$.  These serve as the
hints for clause 36.

Figure \ref{fig:eg4:proof}(C) shows the RUP proof steps required to
delete the input clauses.  Consider the first of these, deleting
input clause $\obar{x}_1 \lor x_3 \lor \obar{x}_4$.  The requirement is to show
that there is no assignment $\assign$ that falsifies this clause, but assigns $\assign(s_{10}) = 1$.
proof proceeds by first assuming that the clause is false, requiring
$\assign(x_1) = 1$, $\assign(x_3) = 0$, and $\assign(x_4) = 1$.  The hints then consist of unit
clauses (e.g., clause 36 asserting that $\alpha(p_{10}) = 1$) or
clauses that cause unit propagation.  Hint clauses 8 and 10 force the
assignments $\assign(p_5) = \assign(p_6) = 0$.  These, plus hint clause 12 forces
$\assign(s_7) = 0$.  This, plus hint clauses 16 and 21 force $\assign(p_8) = \assign(p_9) = 0$, leading
via clause 22 to $\assign(s_10) = 0$.  But this contradicts clause 22,
completing the RUP proof.  The deletion hints for the other input
clauses follow similar patterns---they work from the bottom nodes of
the POG upward, showing that any assignment that falsifies the clause
must assign $s_{10} = 0$.

>>>>

\newcommand{\varset}{X}
\newcommand{\exvarset}{Z}
\newcommand{\dependencyset}{{\cal D}}
\newcommand{\litset}{{\cal L}}
\newcommand{\ring}{{\cal R}}
\newcommand{\dset}{{\cal A}}
\newcommand{\rep}{\textbf{R}}
\newcommand{\radd}{+}
\newcommand{\rmul}{\times}
\newcommand{\addident}{\textbf{0}}
\newcommand{\mulident}{\textbf{1}}
\newcommand{\imply}{\Rightarrow}

@@

\authorrunning{R. E. Bryant, et al} %TODO mandatory. First: Use abbreviated first/middle names. Second (only in severe cases): Use first author plus 'et al.'

@@

\begin{minipage}{0.62\textwidth}
(A)  Input Formula\\[1.2ex]
\begin{tabular}{lll}
\toprule
\makebox[5mm]{ID} & \makebox[15mm]{Clauses} & \\
\midrule
1 & \texttt{-1 3 -4} & \texttt{0} \\
2 & \texttt{-1 -3 4} & \texttt{0} \\
3 & \texttt{3 -4} & \texttt{0}\\
4 & \texttt{1 -3 4} & \texttt{0} \\
5 & \texttt{-1 -2} & \texttt{0} \\
\bottomrule
\end{tabular}
\\[1.8ex]
(C) POG Declaration\\[1.2ex]
\begin{tabular}{llll}
\toprule
\makebox[5mm]{ID} & \multicolumn{2}{l}{CRAT line} & Explanation \\
\midrule
6 & \texttt{p 5 -3 -4} & \texttt{0} & $p_5 = \obar{x}_3 \pand \obar{x}_4$ \\
9 & \texttt{p 6 3 4} & \texttt{0} & $p_6 = x_3 \pand x_4$ \\
12 & \texttt{s 7 5 6 7 10} & \texttt{0} & $s_7 = p_5 \por p_6$ \\
15 & \texttt{p 8 -1 7} & \texttt{0} & $p_8 = \obar{x}_1 \pand s_7$ \\ 
18 & \texttt{p 9 1 -2 7} & \texttt{0} & $p_9 = x_1 \pand \obar{x}_2 \pand s_7$ \\
22 & \texttt{s 10 8 9 16 19} & \texttt{0} & $s_{10} = p_8 \por p_9$ \\
 & \texttt{r 10} && Root $r = s_{10}$\\
\bottomrule
\end{tabular}
\end{minipage}
\begin{minipage}{0.35\textwidth}
(B) POG Representation \\
\input{dd/eg4}
\end{minipage}
%% Break
\\[2.5ex]
\begin{minipage}{0.45\textwidth}
(D) Implicit Clauses\\[1.2ex]
\begin{tabular}{llll}
\toprule
\makebox[5mm]{ID} & \multicolumn{2}{l}{Clauses} & Explanation \\
\midrule
\texttt{6} & \texttt{5 3 4} & \texttt{0} & Define $p_5$ \\
\texttt{7} & \texttt{-5 -3} & \texttt{0} & \\
\texttt{8} & \texttt{-5 -4} & \texttt{0} & \\
\midrule
\texttt{9} & \texttt{6 -3 -4} & \texttt{0} & Define $p_6$ \\
\texttt{10} & \texttt{-6 3} & \texttt{0} & \\
\texttt{11} & \texttt{-6 4} & \texttt{0} & \\
\midrule
\texttt{12} & \texttt{-7 5 6} & \texttt{0} & Define $s_7$ \\
\texttt{13} & \texttt{7 -5} & \texttt{0} & \\
\texttt{14} & \texttt{7 -6} & \texttt{0} & \\
\midrule
\texttt{15} & \texttt{8 1 -7} & \texttt{0} & Define $p_8$ \\
\texttt{16} & \texttt{-8 -1} & \texttt{0} & \\
\texttt{17} & \texttt{-8 7} & \texttt{0} & \\
\midrule
\texttt{18} & \texttt{9 -1 2 -7} & \texttt{0} & Define $p_9$ \\
\texttt{19} & \texttt{-9 1} & \texttt{0} & \\
\texttt{20} & \texttt{-9 -2} & \texttt{0} & \\
\texttt{21} & \texttt{-9 7} & \texttt{0} & \\
\midrule
\texttt{22} & \texttt{-10 8 9} & \texttt{0} & Define $s_{10}$ \\
\texttt{23} & \texttt{10 -8} & \texttt{0} & \\
\texttt{24} & \texttt{10 -9} & \texttt{0} & \\
\bottomrule
\end{tabular}
\end{minipage}
\begin{minipage}{0.49\textwidth}
(E) Assertions\\[1.2ex]
\begin{tabular}{llllll}
\toprule
\makebox[5mm]{ID} & \multicolumn{2}{l}{CRAT line} & \multicolumn{2}{l}{Hints} & Explanation \\
\midrule
\texttt{25} & \texttt{a 5 1 3} & \texttt{0} & \texttt{3 6} & \texttt{0} & $\obar{x}_1 \land \obar{x}_3 \imply p_5$ \\
\texttt{26} & \texttt{a 6 1 -3} & \texttt{0} & \texttt{4 9} & \texttt{0} & $\obar{x}_1 \land x_3 \imply p_6$ \\
\texttt{27} & \texttt{a 3 7 1} & \texttt{0} & \texttt{13 25} & \texttt{0} & $\obar{x}_3 \land \obar{x}_1 \imply s_7$  \\
\texttt{28} & \texttt{a 7 1} & \texttt{0} & \texttt{27 14 26} & \texttt{0} & $\obar{x}_1 \imply s_7$  \\
\texttt{29} & \texttt{a 8 1} & \texttt{0} & \texttt{28 15} & \texttt{0} & $\obar{x}_1 \imply p_8$  \\
\texttt{30} & \texttt{a 5 -1 3} & \texttt{0} & \texttt{1 6} & \texttt{0} & $x_1 \land \obar{x}_3 \imply p_5$ \\
\texttt{31} & \texttt{a 6 -1 -3} & \texttt{0} & \texttt{2 9} & \texttt{0} & $x_1 \land x_3 \imply p_6$ \\
\texttt{32} & \texttt{a 3 7 -1} & \texttt{0} & \texttt{13 30} & \texttt{0} & $\obar{x}_3 \land x_1 \imply s_7$  \\
\texttt{33} & \texttt{a 7 -1} & \texttt{0} & \texttt{32 14 31} & \texttt{0} & $x_1 \imply s_7$  \\
\texttt{34} & \texttt{a 9 -1} & \texttt{0} & \texttt{5 33 18} & \texttt{0} & $x_1 \imply p_9$  \\
\texttt{35} & \texttt{a 1 10} & \texttt{0} & \texttt{23 29} & \texttt{0} & $\obar{x}_1 \imply s_{10}$  \\
\texttt{36} & \texttt{a 10} & \texttt{0} & \texttt{35 24 34} & \texttt{0} & $s_{10}$ \\
\bottomrule
\end{tabular}
\\[1.5ex]
(F) Input Clause Deletions\\[1.2ex]
\begin{tabular}{lllll}
  \toprule
 \multicolumn{3}{l}{CRAT line} & Explanation\\
\midrule
 \texttt{dc 1} & \texttt{36 8 10 12 16 21 22} & \texttt{0} & Delete clause 1 \\
 \texttt{dc 2} & \texttt{36 7 11 12 16 21 22} & \texttt{0} & Delete clause 2 \\
 \texttt{dc 3} & \texttt{36 8 10 12 17 19 22} & \texttt{0} & Delete clause 3 \\
 \texttt{dc 4} & \texttt{36 7 11 12 17 19 22} & \texttt{0} & Delete clause 4 \\
 \texttt{dc 5} & \texttt{36 16 20 22} & \texttt{0} &  Delete clause 5 \\
\bottomrule
\end{tabular}
\end{minipage}
\caption{Example formula (A), its POG representation (B), and its CRAT proof (C)--(F)}
\label{fig:eg4:all}
\end{figure}

Figure \ref{fig:eg4:all} illustrates an example formula and shows how
the CRAT file declares its POG representation.  The input formula (A)
consists of five clauses over variables $x_1$, $x_2$, $x_3$, and
$x_4$.  The generated POG (B) has six nonterminal nodes representing four
conjunctions and two disjunctions.  We name these by the node type (product or sum) and the extension variable.
The first part of the CRAT file
(C) declares these nodes using clause IDs that increment by three or
four, depending on whether the node has two children or three.
The last two nonzero values in the sum declarations are hints providing the required mutual exclusion proof.

We step through a detailed example to provide a better understanding of the CRAT proof framework.
Figure
\ref{fig:eg4:all}(D) shows the defining clauses that are implicitly
defined by the POG operation declarations.  These do not appear in the
CRAT file.  Referring back to the declarations of the sum nodes in
Figure \ref{fig:eg4:all}, we can see that the declaration of node
$s_7$ had clause IDs 7 and 10 as hints.  We can see in Figure
\ref{fig:eg4:all}(A) that these two clauses resolve to the clause
$\obar{p}_5 \lor \obar{p}_6$, showing that the two children of $s_7$
have disjoint models.  Similarly, node $s_{10}$ is declared as having
clause IDs 16 and 19 as hints.  These resolve to the clause
$\obar{p}_8 \lor \obar{p}_9$, showing that the two children of
$s_{10}$ have disjoint models.

Figure \ref{fig:eg4:all}(E) provides the sequence of assertions
leading to unit clause 36, consisting of the literal $s_{10}$.  This clause indicates that $s_{10}$ is implied by the input clauses, i.e.,
any assignment $\assign$
satisfying the input clauses must extend to one with $\assign(s_{10}) = 1$.
Working backward, we can see that
clause 29 indicates that node $p_8$ will be implied by the input
clauses when $\assign(x_1) = 0$, while clause 34 indicates that node $p_9$ will
be implied by the input clauses when $\assign(x_1) = 1$.  These serve as the
hints for clause 36.

Figure \ref{fig:eg4:all}(F) shows the RUP proof steps required to
delete the input clauses.  Consider the first of these, deleting
input clause $\obar{x}_1 \lor x_3 \lor \obar{x}_4$.  The requirement is to show
that there is no assignment $\assign$ that falsifies this clause, but assigns $\assign(s_{10}) = 1$.
proof proceeds by first assuming that the clause is false, requiring
$\assign(x_1) = 1$, $\assign(x_3) = 0$, and $\assign(x_4) = 1$.  The hints then consist of unit
clauses (e.g., clause 36 asserting that $\alpha(p_{10}) = 1$) or
clauses that cause unit propagation.  Hint clauses 8 and 10 force the
assignments $\assign(p_5) = \assign(p_6) = 0$.  These, plus hint clause 12 forces
$\assign(s_7) = 0$.  This, plus hint clauses 16 and 21 force $\assign(p_8) = \assign(p_9) = 0$, leading
via clause 22 to $\assign(s_10) = 0$.  But this contradicts clause 22,
completing the RUP proof.  The deletion hints for the other input
clauses follow similar patterns---they work from the bottom nodes of
the POG upward, showing that any assignment that falsifies the clause
must assign $s_{10} = 0$.
